---
title: "Floods: exploratory data analysis"
author: Zhan Gu
date: 2023 Oct 28
format: html
engine: knitr
---
## Background
Floods are among the most destructive natural disasters. In the United States, natural disaster recovery is managed by (FEMA) Federal Emergency Managment Agency. This report aims to investigate floods in 2020-2021 and try to answer the following questions: 
How dangerous are floods? How expensive? Is there any pattern to the kinds of communities that suffer losses from floods?

## Data sources
[NOAA Storm Events Database](https://www.ncdc.noaa.gov/stormevents/ftp.jsp)
[OpenFEMA Dataset: FEMA Web Disaster Summaries - v1](https://www.fema.gov/openfema-data-page/fema-web-disaster-summaries-v1)
[OpenFEMA Dataset: Disaster Declarations Summaries - v2](https://www.fema.gov/openfema-data-page/disaster-declarations-summaries-v2)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r message=FALSE, warning=FALSE}
library(knitr)  
library(readr)
library(dplyr)
library(kableExtra)
library(tidyverse)
library(ggplot2)
library(stringr)
library(rfema)
library(lubridate)
```
## Overview of FEMA data
Number of Flood Events: There were 14 unique flood events during 2020 to 2021.

Affected States: These flood events impacted 13 different states.

Timeframe of Incidents: The timeframe of the flood events extends from September 17, 2019, to August 21, 2021. This range suggests that some flood events either had prolonged impacts or that their relief and recovery efforts spanned into the 2020-2021 period.

Total Financial Assistance: The total financial assistance provided for these floods amounted to approximately $14.1 billion. This total encompasses various forms of financial assistance, including individual and public assistance, housing assistance, and hazard mitigation grants.

Average Financial Assistance per Event: On average, each flood event received about $1.01 billion in financial assistance.
```{r message=FALSE, warning=FALSE}
# FEMA
declarations_data <- read_csv("Declarations.csv")
disaster_data <- read_csv("Disaster.csv")

# Filter for floods in 2020 and 2021
floods_declarations <- declarations_data %>%
  filter(incidentType == "Flood", fyDeclared %in% c(2020, 2021))

# Merge the data
merged_data <- left_join(floods_declarations, disaster_data, by = "disasterNumber")

# Count the number of flood incidents for each state
state_counts <- merged_data %>% 
  group_by(state) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))

# Count the number of flood incidents for each state
merged_data$totalEconomicLoss <- rowSums(merged_data[,c('totalAmountIhpApproved', 'totalAmountHaApproved', 
                                                       'totalAmountOnaApproved', 'totalObligatedAmountPa', 
                                                       'totalObligatedAmountCatAb', 'totalObligatedAmountCatC2g', 
                                                       'totalObligatedAmountHmgp')], na.rm = TRUE)

state_loss <- merged_data %>% 
  group_by(state) %>% 
  summarise(TotalLoss = sum(totalEconomicLoss, na.rm = TRUE)) %>% 
  arrange(desc(TotalLoss))

# Convert date columns to Date type
date_columns <- c("declarationDate", "incidentBeginDate", "incidentEndDate")
merged_data[date_columns] <- lapply(merged_data[date_columns], ymd)

# Replace missing financial values with 0
financial_columns <- c("totalObligatedAmountPa", "totalObligatedAmountCatAb", 
                       "totalObligatedAmountCatC2g", "totalObligatedAmountHmgp")
merged_data[financial_columns] <- lapply(merged_data[financial_columns], function(x) ifelse(is.na(x), 0, x))

# Delete irrelevant columns
columns_to_keep <- c("disasterNumber", "declarationDate", "incidentBeginDate", "incidentEndDate",
                     "declarationTitle", "state", "designatedArea", 
                     "totalObligatedAmountPa", "totalObligatedAmountCatAb", 
                     "totalObligatedAmountCatC2g", "totalObligatedAmountHmgp")
merged_data <- select(merged_data, one_of(columns_to_keep))

# Summary statistics
summary <- data.frame(
  "Number of Flood Events" = n_distinct(merged_data$disasterNumber),
  "Affected States" = n_distinct(merged_data$state),
  "Timeframe Start" = min(merged_data$incidentBeginDate, na.rm = TRUE),
  "Timeframe End" = max(merged_data$incidentEndDate, na.rm = TRUE),
  "Total Financial Assistance" = sum(sapply(merged_data[financial_columns], sum), na.rm = TRUE)
)
```

```{r message=FALSE, warning=FALSE}
# NOAA
storm2020 <- read.csv("storm2020.csv")
storm2021 <- read.csv("storm2021.csv")
flood2020 <- storm2020 %>% filter(EVENT_TYPE == "Flood")
flood2021 <- storm2021 %>% filter(EVENT_TYPE == "Flood")
combined_data <- bind_rows(flood2020, flood2021)
write.csv(combined_data, "combined_data.csv", row.names = TRUE)

# Convert date columns to Date type
combined_data <- combined_data %>%
  mutate(begin_date = make_date(BEGIN_YEARMONTH / 100, BEGIN_YEARMONTH %% 100, BEGIN_DAY),
         end_date = make_date(END_YEARMONTH / 100, END_YEARMONTH %% 100, END_DAY))

# Delete irrelevant columns
columns_to_select <- c("EPISODE_ID","EVENT_ID","begin_date", "end_date", "STATE", "YEAR", "INJURIES_DIRECT","INJURIES_INDIRECT","DEATHS_DIRECT","DEATHS_INDIRECT","DAMAGE_PROPERTY","DAMAGE_CROPS","FLOOD_CAUSE")
combined_data <- select(combined_data, one_of(columns_to_select))
```


## How dangerous?
```{r message=FALSE, warning=FALSE}
#FEMA
ggplot(state_counts, aes(x = reorder(state, -count), y = count)) + 
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Number of Flood Incidents by State (FEMA)", 
       x = "State", 
       y = "Number of Incidents")

# Counting frequency of floods in each state
frequency_by_state <- combined_data %>%
  count(STATE) %>%
  arrange(desc(n))

# Plotting
ggplot(frequency_by_state, aes(x = reorder(STATE, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Frequency of Floods by State (NOAA)", x = "State", y = "Frequency")

# Calculating duration of each flood event
combined_data <- combined_data %>%
  mutate(duration = as.numeric(difftime(end_date, begin_date, units = "days")))

# Average duration by state
duration_by_state <- combined_data %>%
  group_by(STATE) %>%
  summarize(avg_duration = mean(duration, na.rm = TRUE))

# Plotting
ggplot(duration_by_state, aes(x = reorder(STATE, avg_duration), y = avg_duration)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  labs(title = "Average Duration of Flood Events by State (NOAA)", x = "State", y = "Average Duration (Days)")
```

## How expensive?
```{r message=FALSE, warning=FALSE}
ggplot(state_loss, aes(x = reorder(state, -TotalLoss), y = TotalLoss)) +
  geom_bar(stat="identity", fill="steelblue") +
  theme_minimal() +
  labs(title="Total Economic Loss by State Due to Floods (FEMA)",
       x="State", y="Total Economic Loss")
```

